Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-24 text/ms.tex
Add macro magic numbers to main text

Black swans are statistically improbable events that nonetheless occur ---
often with profound implications. While extremes in the physical
environment,such as monsoons and heat waves, are widely studied and increasing
in magnitude and frequency, it remains unclear the extent to which ecological
populations buffer or suffer from extremes. Here, we develop a probability
model to estimate the degree of heavy-tailedness (presence of black swans) in
ecological process noise. We apply our model to \NPops~time series from around
the world across \NOrders~taxonomic orders and seven classes. We find strong
evidence of black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
(\InsectaRangePerc\%), and fishes(\OsteichthyesRangePerc\%). When they occur,
they tend to be driven by climate and severe winters, cycles of parasites and
predators, and interactions between these elements. They are more frequently
detected for populations with longer time series and lower levels of process
noise. We find little evidence of intrinsic life-history covariates. TODO close
with modelling and policy implications

@@

We obtained abundance time series from the Global Population Dynamics Database
(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
a high-quality subset of populations that were suitable for our analyses, we
filtered the data (Supporting Material). Our filtering removed populations from
less reliable data sources, removed those without sufficient data for our
models, and interpolated some missing values. Our final dataset contained
\NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes
with a median of XX time steps (range of XX--XX) (Table S1,
Fig.~\ref{fig:all-ts}).

@@

We found strong but rare evidence for black-swan dynamics. Defining black-swan
dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black
swans most frequently for birds (X\%) followed by mammals (X\%), insects (X\%),
and fishes (X\%) (Fig.~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans
were taxonomically widespread, occurring for at least one population in X\% of
the taxonomic orders recorded. Accounting for time series length and partially
pooling inference across taxonomic class and order with a multilevel model,
there was stronger evidence for black swans in insect populations than is
apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest
median probability of heavy tails were insect orders --- however, there was
considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Including an
autocorrelation structure in the residuals, modelling population growth rates
without density dependence, or modelling the population dynamics as
Ricker-logistic did not systematically alter our conclusions. Allowing for
a fixed quantity of observation error decreased our estimates of black-swan
dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in X of
XX populations although the majority of $\nu$ estimates remained qualitatively
similar (Fig.~\ref{fig:alt})

Across populations, the probability of observing black swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
length was the strongest covariate of observing black swan dynamics (as was
similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
we were about 15\% TODO more likely to observe heavy tails in a population with
40 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}).
However, the absolute change in probability with increased time series length
was small ($0.X$ vs. $0.X$ in the previous example, Fig.~\ref{fig:correlates}).

@@

We found strong evidence that the multiplicative process noise jumps from time
step to time step were greater than we would expect under a normal distribution
and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We
detected black swans more frequently for longer time series and for populations
with a smaller magnitude of process error, as we might expect. However, we
failed to find strong evidence that black swans were associated with density
dependence, population growth rate, or lifespan. In cases that we verified,
black swans were often a result of the interaction between elements of extreme
climate, predators and parasite cycles, and strong changes in human pressures.
Our empirical results, sensitivity tests, and simulation tests suggest that
estimating the tail shape of process error is a viable method of detecting
heavy-tailed population dynamics and if anything will underestimate the
probability of black swans. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond
to, and are robust to extremes in population dynamics --- particularly as the
frequency and magnitude of climatic extremes increase over the next century
(REF).

>>>>

Black swans are statistically improbable events that nonetheless occur ---
often with profound implications. While extremes in the physical
environment, such as monsoons and heat waves, are widely studied and increasing
in magnitude and frequency, it remains unclear the extent to which ecological
populations buffer or suffer from extremes. Here, we develop a probability
model to estimate the degree of heavy-tailedness (presence of black swans) in
ecological process noise. We apply our model to \NPops~time series from around
the world across \NOrders~taxonomic orders and seven classes. We find strong
evidence of black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
(\InsectaRangePerc\%), and fishes(\OsteichthyesRangePerc\%). When they occur,
they tend to be driven by climate and severe winters, cycles of parasites and
predators, and interactions between these elements. They are more frequently
detected for populations with longer time series and lower levels of process
noise. We find little evidence of intrinsic life-history covariates. TODO close
with modelling and policy implications

@@

We obtained abundance time series from the Global Population Dynamics Database
(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
a high-quality subset of populations that were suitable for our analyses, we
filtered the data (Supporting Material). Our filtering removed populations from
less reliable data sources, removed those without sufficient data for our
models, and interpolated some missing values. Our final dataset contained
\NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes
with a median of \medianTimeSteps~time steps (range of
\minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).

@@

We found strong but rare evidence for black-swan dynamics. Defining black-swan
dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black
swans most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%),
insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs},
\ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring for
at least one population in \POrdersHeavy\% of the taxonomic orders recorded.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a multilevel model, there was stronger evidence
for black swans in insect populations than is apparent in
Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median
probability of heavy tails were insect orders --- however, there was
considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Including an
autocorrelation structure in the residuals, modelling population growth rates
without density dependence, or modelling the population dynamics as
Ricker-logistic did not systematically alter our conclusions. Allowing for
a fixed quantity of observation error decreased our estimates of black-swan
dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in
$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations although the majority of
$\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})

Across populations, the probability of observing black swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
length was the strongest covariate of observing black swan dynamics (as was
similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
in a population with 60 time steps than in one with 30 time steps
(Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with
increased time series length was small ($0.\pHeavyNSixty$ vs.\
$0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

@@

We found strong evidence that the multiplicative process noise jumps from time
step to time step were greater than we would expect under a normal distribution
and a Gompertz or Ricker-logistic model in \overallMinPerc--\overallMaxPerc\%
of populations assessed. We detected black swans more frequently for longer
time series and for populations with a smaller magnitude of process error, as
we might expect. However, we failed to find strong evidence that black swans
were associated with density dependence, population growth rate, or lifespan.
In cases that we verified, black swans were often a result of the interaction
between elements of extreme climate, predators and parasite cycles, and strong
changes in human pressures. Our empirical results, sensitivity tests, and
simulation tests suggest that estimating the tail shape of process error is
a viable method of detecting heavy-tailed population dynamics and if anything
will underestimate the probability of black swans. The presence of black swans
highlights the importance of developing management strategies that detect
quickly, respond to, and are robust to extremes in population dynamics ---
particularly as the frequency and magnitude of climatic extremes increase over
the next century (REF).
